Lord Kelvin's debate over the age of the Earth: an early attempt of interdisciplinary science

### 2023-08-15 ms.tex
intro

\title{Lord Kelvin's debate over the age of the Earth: an early example of failed interdisciplinary science}
\author{Nicholas Wogan}
\maketitle

@@

\section{Kelvin, Lyell, and Darwin and the debate over the age of the Earth}

@@

\section{Lessons from history}

\section{Conclusions}

>>>>

\title{Lord Kelvin's debate over the age of the Earth: an early attempt of interdisciplinary science}
\author{Nicholas Wogan}
\maketitle

@@

Many of the big unsolved problems in science today will require an interdisciplinary approach to solve. For example, the unsolved problem of the origin of life on Earth can only be tackled by drawing upon many different fields of research \citep{Hays_2015}. Geologic expertise is required to understand the earliest fossil records of life, organic chemists are needed to synthesize early life-like replicating molecules, and atmospheric modelers are essential to estimate the plausible conditions on the early Earth under which life began. Another problem that requires broad expertise is projecting the impact of anthropogenic climate change on society. Any forecast will involve atmospheric scientists, glaciologists, economists, and ecologists among many other areas of expertise. Given that important unsolved problems rely on an interdisciplinary approach, it is essential scientists embedded in different departments and academic cultures learn to work together effectively.

One of the earliest attempts of interdisciplinary science is the debate over the age of the Earth in the late 1800s. The physicist Lord Kelvin used a thermodynamic model of Earth's heat flow to estimate Earth was only $\sim 100$ million years old. He brought this calculation to the attention of geologists, who generally assumed the Earth was indefinitely old based on the paradigm of uniformitarianism. A relatively young Earth also troubled Darwin, who worried his theory of biological evolution needed more time than Kelvin allowed. The ensuing debate, involving physics, geology and biology was one of the earliest attempts of interdisciplinary science. However, the scientific discourse did not proceed smoothly. Geologists and physicists talked past each other, and neither discipline was capable of entertaining evidence outside their domain of expertise.

The purpose of this paper is to attempt to shed light on why interdisciplinary science failed during the late 1800s debate over the age of the Earth. Furthermore, I use this historical failure to argue how interdisciplinary science might be best carried out in modern research. To achieve these goals, I first summarize the late 1800s discourse over Earth's age (Section \ref{sec:2}). Next, in Section \ref{sec:3}, I provide possible reasons for why this early attempt of interdisciplinary science was unsuccessful, and what lessons might be drawn from history to improve modern interdisciplinary work. Finally, in Section \ref{sec:4}, I summarize and conclude.

\section{Kelvin, Lyell, and Darwin and the debate over the age of the Earth} \label{sec:2}

@@

\section{Lessons from history} \label{sec:3}

\section{Conclusions} \label{sec:4}
